---
title: "Master script for postfire analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Source functions, get data and plot

First we'll _source()_ (i.e. "run all code in") the scripts with the functions we made. Then we'll set the URL, read in the data with _download.NDVI()_, and plot it with _plot.NDVI()_.

```{r}
## Load required functions by running source() on the individual function files
if(file.exists("01_download.NDVI.R")) source("01_download.NDVI.R")
if(file.exists("02_plot.NDVI.R"))     source("02_plot.NDVI.R")
if(file.exists("03_negexp.R"))        source("03_negexp.R")
## Download NDVI data
URL = "https://raw.githubusercontent.com/jslingsby/BIO3019S_Ecoforecasting/master/data/modisdata.csv"
dat <- download.NDVI(URL)
# Convert "calendar_date" to postfire age in days since fire - assuming the first date in the times eries is the time of the fire 
dat$age <- (as.numeric(dat$calendar_date) - min(as.numeric(dat$calendar_date), na.rm = T))/365.25
## Plot overall NDVI time series
plot.NDVI(dat)
```

<br>

Q1: This plot suggests that Fynbos greenness (NDVI) as observed from satellite saturates with time since fire. Why do you think it saturates rather than increasing linearly with time?

>*Answer 1: There is a finite amount of ground cover that fynbos can occupy, meaning it can support a finite number of plants with its limited resources. After a certain point, there is no space for new plants, as they would be quickly out-competed by already-established plants,  which is why the the plot saturates.* 
<br>


### 2. Fit models using Non-linear Least Squares (NLS)

Now we'll fit the simple and full negative exponential models using Non-linear Least Squares (NLS).

First the simpler model:

```{r}
## Simple model
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)
# fit model
fit_negexp <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda)),
                  data = dat, start = par, trace = F, 
                  control = nls.control(maxiter = 500))
# plot
plot.NDVI(dat = dat, fit = fit_negexp)
```

<br>

And let's look at the model summary with parameter estimates

```{r}
# print model summary
summary(fit_negexp)
```

<br>

Now the full model:

```{r}
## Full model
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)
# fit model
fit_negexpS <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda))
                   + A*sin(2*pi*age + (phi + pi/6*(3 - 1))), 
                   data = dat, start = par, trace = F, 
                   control = nls.control(maxiter = 500))
# plot
plot.NDVI(dat = dat, fit = fit_negexpS)
```


```{r}
# print model summary
summary(fit_negexpS)
```

<br>

Lots more parameters...

Q2: How do the estimates for the common parameters compare?

>*Answer 2:*
From the simple to the full model, alpha and lambda decrease, while gamma increases. It is worth noting that lambda undergoes a greater change than the other 2 parameters.  
<br>

### 3. Compare NLS models using ANOVA

Modelers often want to know which of a set of models are better. One way to do this when comparing nested* models using least squares is using analysis of variance (ANOVA). In this case the `anova()` function will take the model objects as arguments, and return an ANOVA testing whether the full model results in a significant reduction in the residual sum of squares (and thus is better at capturing the data), returning an F-statistic, Degrees of Freedom (the difference in the number of parameters between the models) and p-value.

*i.e. one model is a subset of the other, as in our case

```{r}
anova(fit_negexp, fit_negexpS)
```

<br>

Q3: Which model is better?

>*Answer 3:*
Model 2 is better as it has a lower residual sum of squares, showing that it captures data better.
<br>

Q4: How many degrees of freedom are there in this ANOVA and why (i.e. what are they)?

>*Answer 4:*
In an anova model there are many degrees of freedom. In this case, there are 3 types:
1. For the simple model: There are 3 parameters (lambda, alpha and gamma). Therefore 135 (Residual dfs) + 3  parameters = 138 degrees of freedom. 
2. For the full model: There are 5 parameters (lambda, alpha, gamma, A and phi). Therefore 133 (Residual dfs) + 5  parameters = 138 degrees of freedom. 
Thus there are a total 138 degrees of freedom for both.
3. The degrees of freedom are the difference in the number of parameters between the models: In this case it is 5 - 3 = 2

The degrees of freedom are the number of observations - 1. They can be thought of as the number of independent deviations around a mean.
<br>

### 4. Fit models using Maximum Likelihood Estimation (MLE)

First let's fit the simpler model:

```{r}
## Fit the simpler model using MLE
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)
# fit model
fit_negexpMLE <- fit.negexp.MLE(dat, par)
# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexp(fit_negexpMLE$par,dat$age), col = 'skyblue', lwd = 3)
```


```{r}
fit_negexpMLE
```

<br>

Then the full model:

```{r}
## Fit the full model using MLE
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)
# fit model
fit_negexpMLES <- fit.negexpS.MLE(dat, par)
# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexpS(fit_negexpMLES$par,dat$age), col = 'skyblue', lwd = 3)
```

```{r}
fit_negexpMLES
```

<br>

### 5. Compare MLE models using Akaike's information criterion (AIC)

Note that we can't compare our MLE models using ANOVA because our custom functions do not return full model fits like the `nls()` function - only the parameter estimates, negative log-likelihoods and a few other diagnostics.

Another way to compare models (and probably the most common) is using the Akaike information criterion (AIC), which is an estimator of prediction error (i.e. relative quality) of statistical models for a given set of data. 

The formula for the Akaike information criterion is:

$AIC = 2K -2(ln(L))$

Where:

- $k$ = the number of estimated parameters in the model
- $L$ = maximum value of the likelihood function for the model

Since we have our negative log likelihoods (i.e. $-ln(L)$ in the formula above), we can calculate the AICs and compare them.

```{r}
AIC_simple = 6 + 2*fit_negexpMLE$value
AIC_simple
AIC_full = 6 + 2*fit_negexpMLES$value
AIC_full
```

<br>

When comparing models, the lower the AIC the better, and in general a difference in AIC of 3 or more is analagous to the models being significantly different at an $\alpha$ of $p < 0.05$.

```{r}
AIC_simple - AIC_full
```

<br>

Q5: Is there a preferred model and if so, which one?

>*Answer 5:*
Yes, it is the full model, because it has a drastically lower AIC value. The difference between them in AIC value is much greater than 3, showing that they are extremely different and the full model is much better.
<br>

The nice thing about AIC is that the models you compare do not have to be nested like they do for ANOVA, as long as the data are the same. There are a few other constraints however... 

Here are the AIC scores for our pair of NLS models:

```{r}
AIC(fit_negexp, fit_negexpS)
```

<br>

You'll notice that these are completely different to the AICs for the MLE models...

Q6: Why is it not okay to compare the AIC of these NLS models with the AIC of the MLE models? Hint: type `?AIC` into the R console and do some reading.

>*Answer 6:*
We cannot because the MLE is done using log likelihood while the NLS is done using least squares. Any model not fitted by maximum likelihood cannot have its AIC values compared to one done with it. 
<br>